---
title: "transcript_vs_translation_within_KEGG_groups"
author: "Chao Di, dic@email.chop.edu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc_depth: 1
    number_sections: true
---

```{r setup, include=FALSE, echo = FALSE, cache = FALSE}
# replace with path where you want the results be
knitr::opts_knit$set(root.dir="/Users/dic/Documents/Will Bailis lab/GSE120762_riboseq/KEGG_groups_mouse")
```

## Abstract {-}
This is to analyze the correlation/association of transcription and translation for genes within the same KEGG functional groups (modules/pathways)

```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE}
library(dplyr)
library(tidyr)
library(tibble)
library(data.table)
library(reshape2)
library(ggplot2)
library(ggnewscale)
```

### Prepare tables of KEGG modules/pathways with associated genes {-}
Results from shell script: link_pathway-module-gene.sh
The .sh script was complied for retrieving all mouse KEGG modules or pathways and find associated genes with annotations. KEGG API was used in the script. R package `KEGGREST` should be sufficient for the same work.

```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE}
# read module/pathway-gene association tables
setwd("/Users/dic/Documents/Will Bailis lab/GSE120762_riboseq/KEGG_groups_mouse")
df_mod_gene <- read.table("mouse_modulesAnno_with_mmuID_genename.txt", header=TRUE, sep="\t", quote="")
head(df_mod_gene)
mod_nGenes <- read.table("mouse_modules_nGenes.txt",  header=TRUE, sep="\t") 
head(mod_nGenes)

df_path_gene <- read.table("mouse_pathwaysAnno_with_mmuID_genename.txt", header=TRUE, sep="\t", quote="")
df_path_gene$pathAnno = gsub(" - Mus musculus (mouse)","",df_path_gene$pathAnno, fixed = TRUE)
head(df_path_gene)
path_nGenes <- read.table("mouse_pathways_nGenes.txt",  header=TRUE, sep="\t") 
path_nGenes$pathAnno = gsub(" - Mus musculus (mouse)","",path_nGenes$pathAnno, fixed = TRUE)
head(path_nGenes)
```

### Correlation analysis -- within modules {-}
* Add gene expression (read counts from RNA-seq) and gene translation (read counts from ribo-seq) values
* Calculate Correlation Coefficients (Pearson or Spearman) within each module
* Report the r and p values

```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE}

setwd("/Users/dic/Documents/Will Bailis lab/GSE120762_riboseq/KEGG_groups_mouse")
# read RNA-seq expression values (normalized and log2 transformed) from DESeq2
gene_exp <- read.table("../../GSE120762_rnaseq/gene_exp_normalized_transformed.tsv", header=TRUE, sep="\t")

# read RPF normalized read counts from ribo-seq
RPF_RPKM <- read.table("../RPF_rpkm_table.tsv", header=TRUE, sep="\t")

# merge the data above
rna_ribo_exp <- merge(RPF_RPKM, gene_exp, by="GeneID")
colnames(rna_ribo_exp) = c("GeneID","Length", "WT_ribo1", "LPS_ribo1", "WT_ribo2","LPS_ribo2","WT_RNA1","WT_RNA2","WT_RNA3","LPS_RNA1","LPS_RNA2","LPS_RNA3")

# get mean of the replicates
rna_ribo_exp_ave <- rna_ribo_exp %>% rowwise %>% dplyr::mutate(WT_ribo = mean(WT_ribo1, WT_ribo2))
rna_ribo_exp_ave <- rna_ribo_exp_ave %>% rowwise %>% dplyr::mutate(LPS_ribo = mean(LPS_ribo1, LPS_ribo2))
rna_ribo_exp_ave <- rna_ribo_exp_ave %>% rowwise %>% dplyr::mutate(WT_RNA = mean(WT_RNA1, WT_RNA2, WT_RNA3))
rna_ribo_exp_ave <- rna_ribo_exp_ave %>% rowwise %>% dplyr::mutate(LPS_RNA = mean(LPS_RNA1, LPS_RNA2, LPS_RNA3))
rna_ribo_exp_ave <- rna_ribo_exp_ave[,c(1,13:16)]
colnames(rna_ribo_exp_ave)[1] <- "geneName"

# only do for modules with 3 or more genes (CI can be only calculated when sample size>=4)
df_mod_gene = df_mod_gene %>% tidyr::separate(geneAnno,c("geneName","geneAnno"), sep="; ")
df_mod_gene_values = left_join(df_mod_gene, rna_ribo_exp_ave, by="geneName")
# remove genes with NA values, and after that remove modules with less than 3 genes remained
df_mod_gene_values_filtered =df_mod_gene_values %>% drop_na() %>% group_by(modID) %>% filter(n() >= 3)
# add chnanges
df_mod_gene_values_filtered$Change_ribo=df_mod_gene_values_filtered$LPS_ribo - df_mod_gene_values_filtered$WT_ribo
df_mod_gene_values_filtered$Change_RNA=df_mod_gene_values_filtered$LPS_RNA - df_mod_gene_values_filtered$WT_RNA

corr_tb <- df_mod_gene_values_filtered %>% 
  group_by(modID, modAnno) %>% 
  summarise(WT_corr=as.numeric(cor.test(WT_ribo, WT_RNA)$estimate), WT_p=cor.test(WT_ribo, WT_RNA)$p.value, 
            LPS_corr=as.numeric(cor.test(LPS_ribo, LPS_RNA)$estimate), LPS_p=cor.test(LPS_ribo, LPS_RNA)$p.value, 
            Change_corr=as.numeric(cor.test(Change_ribo, Change_RNA)$estimate), Change_p=cor.test(Change_ribo, Change_RNA)$p.value,
            RNA_corr=as.numeric(cor.test(WT_RNA, LPS_RNA)$estimate), RNA_p=cor.test(WT_RNA, LPS_RNA)$p.value,
            ribo_corr=as.numeric(cor.test(WT_ribo, LPS_ribo)$estimate), ribo_p=cor.test(WT_ribo, LPS_ribo)$p.value,
            nGenes=n()) 

# keep only modules with at least one significant correlation and filter only corr difference is more than 0.1
corr_RNA_ribo <- corr_tb %>% filter(WT_p<=0.05 | LPS_p<=0.05) %>% filter(abs(WT_corr - LPS_corr) > 0.1)
corr_WT_LPS <- corr_tb %>% filter(RNA_p<=0.05 | ribo_p<=0.05) %>% filter(abs(RNA_corr - ribo_corr) > 0.1)

```

#### 1. Plot corr and p-value of RNA vs ribo for both WT and LPS, and compare how the values change between WT and LPS {-}
```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE, fig.height=5, fig.width=12}
# reformat the data table
mm=melt.data.table(setDT(corr_RNA_ribo[,c(2:6,13)]), id=c("modAnno","nGenes"), measure=patterns("_corr", "_p"), value.name=c("corr","pvalue"))
mm$variable <- recode(mm$variable, "1"="WT", "2"="LPS")

# visualize the results
p <- ggplot(mm, aes(reorder(modAnno, corr), corr)) +
     geom_point(data=mm[mm$variable=="WT"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("WT",low="dodgerblue",high="grey") +
     scale_color_manual("WT", values=c("p<=0.05"="dodgerblue", "p>0.05"="grey")) +
     new_scale_color() + 
     geom_point(data=mm[mm$variable=="LPS"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("LPS", low="red", high="grey") +
     scale_color_manual("LPS", values=c("p<=0.05"="red", "p>0.05"="grey")) +
     coord_flip() +
     labs(x="Module", y="corr", title= "RNA-ribo correlation comparison between WT and LPS (KEGG modules)", color="p-value", size="nGenes") + 
     theme_minimal()
     #guides(nGenes = guide_legend(order=3))
print(p)
```

#### 2. Plot corr and p-value of WT vs LPS for both RNA-seq and ribo-seq, and compare how the values change between RNA transcription and ribo translation {-}
```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE, fig.height=8, fig.width=12}
# reformat the data table
mm=melt.data.table(setDT(corr_WT_LPS[,c(2,9:13)]), id=c("modAnno","nGenes"), measure=patterns("_corr", "_p"), value.name=c("corr","pvalue"))
mm$variable <- recode(mm$variable, "1"="RNA", "2"="ribo")

# visualize the results
p <- ggplot(mm, aes(reorder(modAnno, corr), corr)) +
     geom_point(data=mm[mm$variable=="RNA"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("RNA",low="dodgerblue",high="grey") +
     scale_color_manual("RNA", values=c("p<=0.05"="dodgerblue", "p>0.05"="grey")) +
     new_scale_color() + 
     geom_point(data=mm[mm$variable=="ribo"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("ribo", low="red", high="grey") +
     scale_color_manual("ribo", values=c("p<=0.05"="red", "p>0.05"="grey")) +
     coord_flip() +
     labs(x="Module", y="corr", title= "WT-LPS comparison between RNA and ribo (KEGG modules)", color="p-value", size="nGenes") + 
     theme_minimal()
     #guides(nGenes = guide_legend(order=3))
print(p)
```

### Correlation analysis -- within pathways {-}
```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE}

# only do for pathways with 3 or more genes (CI can be only calculated when sample size>=4)
df_path_gene = df_path_gene %>% tidyr::separate(geneAnno,c("geneName","geneAnno"), sep="; ")
df_path_gene_values = left_join(df_path_gene, rna_ribo_exp_ave, by="geneName")
# remove genes with NA values, and after that remove modules with less than 3 genes remained
df_path_gene_values_filtered <- df_path_gene_values %>% drop_na() %>% group_by(pathID) %>% filter(n() >= 3)
# add chnanges
df_path_gene_values_filtered$Change_ribo=df_path_gene_values_filtered$LPS_ribo - df_path_gene_values_filtered$WT_ribo
df_path_gene_values_filtered$Change_RNA=df_path_gene_values_filtered$LPS_RNA - df_path_gene_values_filtered$WT_RNA

corr_tb <- df_path_gene_values_filtered %>% 
  group_by(pathID, pathAnno) %>% 
  summarise(WT_corr=as.numeric(cor.test(WT_ribo, WT_RNA)$estimate), WT_p=cor.test(WT_ribo, WT_RNA)$p.value, 
            LPS_corr=as.numeric(cor.test(LPS_ribo, LPS_RNA)$estimate), LPS_p=cor.test(LPS_ribo, LPS_RNA)$p.value, 
            Change_corr=as.numeric(cor.test(Change_ribo, Change_RNA)$estimate), Change_p=cor.test(Change_ribo, Change_RNA)$p.value,
            RNA_corr=as.numeric(cor.test(WT_RNA, LPS_RNA)$estimate), RNA_p=cor.test(WT_RNA, LPS_RNA)$p.value,
            ribo_corr=as.numeric(cor.test(WT_ribo, LPS_ribo)$estimate), ribo_p=cor.test(WT_ribo, LPS_ribo)$p.value,
            nGenes=n()) 

# keep only modules with at least one significant correlation and filter only corr difference is more than 0.2 
corr_RNA_ribo <- corr_tb %>% filter(WT_p<=0.05 | LPS_p<=0.05) %>% filter(abs(WT_corr - LPS_corr) > 0.2)
corr_WT_LPS <- corr_tb %>% filter(RNA_p<=0.05 | ribo_p<=0.05) %>% filter(abs(RNA_corr - ribo_corr) > 0.2)

```

#### 1. Plot corr and p-value of RNA vs ribo for both WT and LPS, and compare how the values change between WT and LPS {-}
```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE, fig.height=4, fig.width=12}
# reformat the data table
mm=melt.data.table(setDT(corr_RNA_ribo[,c(2:6,13)]), id=c("pathAnno","nGenes"), measure=patterns("_corr", "_p"), value.name=c("corr","pvalue"))
mm$variable <- recode(mm$variable, "1"="WT", "2"="LPS")

# visualize the results
p <- ggplot(mm, aes(reorder(pathAnno, corr), corr)) +
     geom_point(data=mm[mm$variable=="WT"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("WT",low="dodgerblue",high="grey") +
     scale_color_manual("WT", values=c("p<=0.05"="dodgerblue", "p>0.05"="grey")) +
     new_scale_color() + 
     geom_point(data=mm[mm$variable=="LPS"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("LPS", low="red", high="grey") +
     scale_color_manual("LPS", values=c("p<=0.05"="red", "p>0.05"="grey")) +
     coord_flip() +
     labs(x="Pathway", y="corr", title= "RNA-ribo correlation comparison between WT and LPS (KEGG pathways)", color="p-value", size="nGenes") + 
     theme_minimal()
     #guides(nGenes = guide_legend(order=3))
print(p)
```

#### 2. Plot corr and p-value of WT vs LPS for both RNA-seq and ribo-seq, and compare how the values change between RNA transcription and ribo translation {-}
```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE, fig.height=18, fig.width=12}
# reformat the data table
mm=melt.data.table(setDT(corr_WT_LPS[,c(2,9:13)]), id=c("pathAnno","nGenes"), measure=patterns("_corr", "_p"), value.name=c("corr","pvalue"))
mm$variable <- recode(mm$variable, "1"="RNA", "2"="ribo")

# visualize the results
p <- ggplot(mm, aes(reorder(pathAnno, corr), corr)) +
     geom_point(data=mm[mm$variable=="RNA"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("RNA",low="dodgerblue",high="grey") +
     scale_color_manual("RNA", values=c("p<=0.05"="dodgerblue", "p>0.05"="grey")) +
     new_scale_color() + 
     geom_point(data=mm[mm$variable=="ribo"], aes(size=nGenes, color=ifelse(pvalue<=0.05, "p<=0.05", "p>0.05"))) +
     #scale_color_gradient("ribo", low="red", high="grey") +
     scale_color_manual("ribo", values=c("p<=0.05"="red", "p>0.05"="grey")) +
     coord_flip() +
     labs(x="Pathway", y="corr", title= "WT-LPS comparison between RNA and ribo (KEGG pathways)", color="p-value", size="nGenes") + 
     theme_minimal()
     #guides(nGenes = guide_legend(order=3))
print(p)
```

### Dissecting groups of genes with translation changes


