---
title: "J2seq analysis, focusing on Ad5"
author: "Chao Di, dic@email.chop.edu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc_depth: 1
    number_sections: true
---

```{r setup, include=FALSE, echo = FALSE, cache = FALSE}
# replace with path where you want the results to be
knitr::opts_knit$set(root.dir="/Users/dic/Documents/Weitzman lab/J2seq")
```

## Abstract {-}
This is the analysis of J2seq data, aim to uncover the role of dsRNA produced by Ad5 or Ad5 mutant(E4) in human cells.
Human cell line (A549) was infected with Adenovirus, either wildtype serotype number 5 Adenovirus (Ad5) or a mutant of Ad5 that is lacking a region of viral proteins in the E4 region (Delta-E4, often just shortened to E4). Post infection the virus has "taken over" the infected cell so completely that more than half of all mRNA is viral. There are 80 million total mappable reads across all 8 libraries vs the about 400 million we expected.

```{r include=TRUE,echo=FALSE,message = FALSE, warning = FALSE}
setwd("/Users/dic/Documents/Weitzman lab/J2seq")
library(dplyr)
library(ggplot2)
library(pheatmap)
library(VennDiagram)
library(DESeq2)
library(edgeR)
library(EnhancedVolcano)
library(fgsea)
library(gprofiler2)
library(tibble)
library(data.table)
library(gprofiler2)
# library(clusterProfiler)
```


### Using DESeq2 for the analysis {-}
Get RNA expression (normalized, transformed using DESeq2) values from the RNA-seq, and do DE
```{r, include=TRUE,echo=FALSE,message = FALSE, warning=FALSE}

## load data, featureCount reads
#cts_human <- read.table("J2seq_featureCount.tsv", header=TRUE, check.names=FALSE)
cts_ad5 <- read.table("J2seq_featureCount_ad5_fwd.tsv", header=TRUE, check.names=FALSE)
cts_ad5_antisense <- read.table("J2seq_featureCount_ad5_rev.tsv", header=TRUE, check.names=FALSE)
rownames(cts_ad5_antisense) <- paste0("anti-", rownames(cts_ad5_antisense))
cts_ad5_segs <- read.table("J2seq_featureCount_ad5_segments.tsv", header=TRUE, check.names=FALSE)
cts <- rbind(cts_ad5, cts_ad5_antisense, cts_ad5_segs)

colnames(cts) <- colnames(cts) %>% gsub(".merged.dedup.sorted.bam", "", .) %>% gsub("\\.", "_", .)

# colData and countData must have the same sample order, but this is ensured by the way we create the count matrix
coldata <- read.table("sample_contrast.tsv",header=TRUE, check.names=FALSE)
coldata$sample <- coldata$sample %>% gsub("-", "_", .)
coldata$lib_type <- coldata$lib_type %>% gsub("-","_", .)
cat("-- This is the experiment design matrix: \n")
coldata

rownames(coldata)=coldata$sample
coldata$phenotype <- factor(coldata$phenotype)
coldata$lib_type <- factor(coldata$lib_type)

## construct DESeq dataset with contrast design
cat("-- Use design = ~ phenotype + lib_type + phenotype:lib_type to build the Negative Binomial GLM (generalized linear model)...\n")
dds <- DESeqDataSetFromMatrix(countData=cts, colData=coldata, design = ~ phenotype + lib_type + phenotype:lib_type)
#dds <- DESeqDataSetFromMatrix(countData=cts, colData=coldata, design = ~ lib_type + phenotype + lib_type:phenotype)
dds$phenotype = relevel(dds$phenotype, "wt")
dds$lib_type = relevel(dds$lib_type, "input")

# remove uninformative columns
# dds <- dds[rowSums(counts(dds)) > 1, ]
# rows with at least one gene >=10
dds <- dds[rowSums(counts(dds) >= 10) > 0, ]

## main function: GLM modeling and differential test
dds <- DESeq(dds, fitType = "local")
cat("-- Using wt and input as the contrast reference, these are the coefficients:\n ")
resultsNames(dds)

## ---- raw count normalization and count transformation, log2 scale, either rlog or vst
# RNA-seq
vsd <- varianceStabilizingTransformation(dds, blind=FALSE)
norm_cts <- assay(vsd)
# cat("Normalized read counts by DESeq2:\n")
# head(norm_cts)
norm_cts <- as.data.frame(cbind(GeneID=rownames(norm_cts), norm_cts))
cat("Do reads normalization and transformation (log2 scale) using rlog or vst, and output a table to *deseq2_cts_normalized_transformed.tsv*. \n")
write.table(norm_cts, file="deseq2_ad5_cts_normalized_transformed.tsv", sep="\t", quote=FALSE, row.names=FALSE)

```

### PCA plot {-}
The PCA plot shows the samples in the 2D plane spanned by their first two principal components. This type of plot is useful for visualizing the overall effect of experimental covariates and batch effects.

```{r, include=TRUE,echo=FALSE,message = FALSE}
## visualizations ##
# pca plot
# pdf("pca-plot.pdf")
pcaData <- plotPCA(vsd, intgroup=c("lib_type","phenotype"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))
ggplot(pcaData, aes(PC1, PC2, color=phenotype, shape=lib_type)) +
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
# dev.off()
```

### Heatmap for expression matrix {-}
Heatmap showing the expressions of the most variable genes (top 50 here),  also do clustering between samples. This gives us the information of which samples and which genes are close to each other. Can also applied to differential genes in later analysis.

```{r, include=TRUE,fig.height=8, fig.width=6,echo=FALSE,message = FALSE}
# heatmap of the count matrix
select <- head(order(-rowVars(assay(vsd))),50) # most variable genes
pdata <- assay(vsd)[select,]
df <- as.data.frame(colData(dds)[,c("phenotype","lib_type")])
rownames(df) <- rownames(colData(dds))
colnames(df) <- c("phenotype","lib_type")
# pdf("heatmap_expression_matrix.pdf")
pheatmap(pdata, cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, annotation_col=df)
# dev.off()
```


### Differential expression between J2-RIP vs input {-}

```{r, include=TRUE,echo=FALSE,message = FALSE, warning=FALSE}

# ---- the RIP effect for wt (reference)
#res_wt <- results(dds, name="lib_type_J2_RIP_vs_input")
res_wt <- results(dds, contrast = c("lib_type", "J2_RIP", "input"))
res_wt_up <- res_wt[which(res_wt$padj<0.05 & res_wt$log2FoldChange>1),]
cat("-- J2-RIP vs input for wt, # of enriched genes (padj<0.05, log2FC>1): \n")
dim(res_wt_up)[1]
res_wt_up
res_wt_up <- as.data.frame(cbind(gene_name=rownames(res_wt_up), res_wt_up))
write.table(res_wt_up, file="deseq2_J2RIP_enriched_genes_wt.tsv", sep="\t", quote=FALSE, row.names=FALSE)


# ---- the RIP effect for e4 
res_e4 <- results(dds, list(c("lib_type_J2_RIP_vs_input", "phenotypee4.lib_typeJ2_RIP")))
res_e4_up <- res_e4[which(res_e4$padj<0.05 & res_e4$log2FoldChange>1),]
cat("-- J2-RIP vs input for e4, # of enriched genes (padj<0.05, log2FC>1):  \n")
dim(res_e4_up)[1]
res_e4_up
res_e4_up <- as.data.frame(cbind(gene_name=rownames(res_e4_up), res_e4_up))
write.table(res_e4_up, file="deseq2_J2RIP_enriched_genes_e4.tsv", sep="\t", quote=FALSE, row.names=FALSE)


# ---- intersect of wt and e4 enriched genes
cat("-- Venn diagram showing the overlap genes between wt and e4: \n")
x=list(wt=rownames(res_wt_up), e4=rownames(res_e4_up))
p <- venn.diagram(x, filename=NULL,  lwd = 2,
        lty = 'blank', fill = c("red","green"),
        cex = 2, # numbers
        cat.cex = 3, cat.fontface = "bold", at.default.pos = "outer", # labels
        print.mode=c("raw","percent"))
grid.newpage()
grid.draw(p)

core_genes = merge(as.data.frame(res_wt_up), as.data.frame(res_e4_up), by='gene_name')
colnames(core_genes) <- colnames(core_genes) %>% gsub(".x", ".wt", .) %>% gsub(".y", ".e4", .) 
cat("Generate a table *deseq2_J2RIP_enriched_genes_wt_and_e4.tsv*: \n")
colnames(core_genes)[1] <- "gene_name" 
write.table(core_genes, file="deseq2_J2RIP_enriched_genes_wt_and_e4.tsv", sep="\t", quote=FALSE, row.names=FALSE)
head(core_genes)



```
