---
title: "Onlink data analysis"
author: "Chao Di, dic@email.chop.edu"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc_depth: 1
    number_sections: yes
  pdf_document:
    toc_depth: '1'
---

```{r setup, include=FALSE, echo = FALSE, cache = FALSE, warning=FALSE}
# replace with path where you want the results be
knitr::opts_knit$set(root.dir="/Users/dic/Documents/Sarah Henrickson lab/Olink/")
```

## Abstract {-}
Olink data analysis from the Normalized Protein eXpression (NPX) using R package `OlinkAnalyze`.
NPX is Olink’s relative protein quantification unit on a log2 scale. NPX values from Olink Explore are calculated from the number of matched counts (= i.e. the number of  sequence reads for each specific combination of sample and assay) from the NovaSeq 6000 or NextSeq 2000 run.

```{r, include=FALSE}
library(readxl)
library(tidyverse)
library(dplyr)
library(OlinkAnalyze)
library(ggplot2)
```

### Preprocessing {-}

```{r}
## load data
data <- read.xlsx("20211163_Henrickson_with_young_HC_NPX_2021-07-29.xlsx")
SampleID_Group <- read.table("SampleID_Group.txt",header = T)

data <- data %>% left_join(SampleID_Group)
## Randomize samples on plate
# olink_plate_randomizer(data, SubjectColumn ="SampleID", iterations=1000, seed=111)

```

### QC {-}

* Scatterplot for QC (olink_qc_plot)
The olink_qc_plot function generates a facet plot per Panel using ggplot and ggplot2::geom_point and stats::IQR plotting IQR vs. median for all samples. This is a good first check to find out if any samples have a tendency to be classified as outliers. Horizontal dashed lines indicate +/-3 standard deviations from the mean IQR. Vertical dashed lines indicate +/-3 standard deviations from the mean sample median.

```{r}

data %>% filter(!str_detect(SampleID, 'CONTROL_SAMPLE')) %>% olink_qc_plot(color_g = "QC_Warning")   
```

* Boxplots for QC (olink_dist_plot)
The olink_dist_plot function generates boxplots of NPX values for each sample, faceted by Olink panel. This is mainly used as an initial QC step to identify potential outliers. olink_dist_plot uses the functions ggplot and geom_boxplot of the R library ggplot2.

```{r}
data %>% # For this example only plotting one panel.
   filter(!str_detect(SampleID, 'CONTROL_SAMPLE')) %>% 
  olink_dist_plot() +
  theme(axis.text.x = element_text(angle=45, hjust=1, size=6)) # Due to the number of samples one can remove the text or rotate it


```

### PCA plot {-}
```{r}

data %>% filter(!str_detect(SampleID, 'CONTROL_SAMPLE')) %>% 
         olink_pca_plot(., color_g = "QC_Warning")  
```


### Group comparison {-}
Boxplots for outcomes (olink_boxplot). The olink_boxplot function is used to generate boxplots of NPX values stratified on a variable for a given list of proteins. olink_boxplot uses the functions ggplot and geom_boxplot of the R library ggplot2.
```{r}

p <- data %>% olink_boxplot(variable = "Group",
                olinkid_list = c("OID20486","OID20487"))
p[[1]]
```

### Statistical analysis {-}
The olink_volcano_plot function generates a volcano plot using results from the olink_ttest function using the function ggplot and geom_point of the R library ggplot2. The estimated difference is shown in the x-axis and -log10(p-value) in the y-axis. A horizontal dotted line indicates p-value=0.05. Dots are colored based on Benjamini-Hochberg adjusted p-value cutoff 0.05 and can optionally be annotated by OlinkID.

* perform t-test for STAT1GOF vs Healthy_Controls

```{r}
# perform t-test for STAT1GOF
data1 = data %>% filter(Group %in% c('Healthy_Controls_young', 'STAT1GOF')) 
data1$Group <- factor(data1$Group, levels=c("STAT1GOF", "Healthy_Controls_young"))
ttest_results <- olink_ttest(df = data1, variable = 'Group')
write.table(as.data.frame(ttest_results), file="STAT1GOF_vs_HC_ttest_results.tsv", sep="\t", quote=F, row.names = F)
# select names of proteins to show
top_10_name <- ttest_results %>%
  slice_head(n = 10) %>%
  pull(OlinkID)
markers <- ttest_results %>% filter(Assay %in% c("IL6", "IL10", "PDCD1")) %>% pull(OlinkID)
labels <- c(top_10_name, markers)
# volcano plot
options(ggrepel.max.overlaps = Inf)
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = 'Variant_vs_Healthy',
                   olinkid_list = labels)
cat("top 10 proteins and IL-6, IL-10 and PDCD1 are labeld")
```

* perform t-test for STAT3GOF vs Healthy_Controls
```{r}
# perform t-test for STAT3GOF
data1 = data %>% filter(Group %in% c('Healthy_Controls_young', 'STAT3GOF'))
data1$Group <- factor(data1$Group, levels=c("STAT3GOF", "Healthy_Controls_young"))
ttest_results <- olink_ttest(df = data1, variable = 'Group')
write.table(as.data.frame(ttest_results), file="STAT3GOF_vs_HC_ttest_results.tsv", sep="\t", quote=F, row.names = F)
# select names of proteins to show
top_10_name <- ttest_results %>%
  slice_head(n = 10) %>%
  pull(OlinkID)
markers <- ttest_results %>% filter(Assay %in% c("IL6", "IL10", "PDCD1")) %>% pull(OlinkID)
labels <- c(top_10_name, markers)
# volcano plot
options(ggrepel.max.overlaps = Inf)
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = 'Variant_vs_Healthy',
                   olinkid_list = labels)
cat("top 10 proteins and IL-6, IL-10 and PDCD1 are labeld")

```

* perform t-test for STAT3LOF vs Healthy_Controls
```{r}
# perform t-test for STAT3LOF
data1 = data %>% filter(Group %in% c('Healthy_Controls_young', 'STAT3LOF')) 
data1$Group <- factor(data1$Group, levels=c("STAT3LOF", "Healthy_Controls_young"))

ttest_results <- olink_ttest(df = data1, variable = 'Group')
write.table(as.data.frame(ttest_results), file="STAT3LOF_vs_HC_ttest_results.tsv", sep="\t", quote=F, row.names = F)
# select names of proteins to show
top_10_name <- ttest_results %>%
  slice_head(n = 10) %>%
  pull(OlinkID)
markers <- ttest_results %>% filter(Assay %in% c("IL6", "IL10", "PDCD1")) %>% pull(OlinkID)
labels <- c(top_10_name, markers)
# volcano plot
options(ggrepel.max.overlaps = Inf)
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = 'Variant_vs_Healthy',
                   olinkid_list = labels)
cat("top 10 proteins and IL-6, IL-10 and PDCD1 are labeld")

```

* perform t-test for CTLA4 vs Healthy_Controls
```{r}
# perform t-test for CTLA4
data1 = data %>% filter(Group %in% c('Healthy_Controls_young', 'CTLA4'))
data1$Group <- factor(data1$Group, levels=c("CTLA4", "Healthy_Controls_young"))

ttest_results <- olink_ttest(df = data1, variable = 'Group')
write.table(as.data.frame(ttest_results), file="CTLA4_vs_HC_ttest_results.tsv", sep="\t", quote=F, row.names = F)
# select names of proteins to show
top_10_name <- ttest_results %>%
  slice_head(n = 10) %>%
  pull(OlinkID)
markers <- ttest_results %>% filter(Assay %in% c("IL6", "IL10", "PDCD1")) %>% pull(OlinkID)
labels <- c(top_10_name, markers)
# volcano plot
options(ggrepel.max.overlaps = Inf)
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = 'Variant_vs_Healthy',
                   olinkid_list = labels)
cat("top 10 proteins and IL-6, IL-10 and PDCD1 are labeld")

```

* perform t-test for PTCRA vs Healthy_Controls
```{r}
# perform t-test for PTCRA
data1 = data %>% filter(Group %in% c('Healthy_Controls_young', 'PTCRA')) 
data1$Group <- factor(data1$Group, levels=c("PTCRA", "Healthy_Controls_young"))


ttest_results <- olink_ttest(df = data1, variable = 'Group')

write.table(as.data.frame(ttest_results), file="PTCRA_vs_HC_ttest_results.tsv", sep="\t", quote=F, row.names = F)
# select names of proteins to show
top_10_name <- ttest_results %>%
  slice_head(n = 10) %>%
  pull(OlinkID)
markers <- ttest_results %>% filter(Assay %in% c("IL6", "IL10", "PDCD1")) %>% pull(OlinkID)
labels <- c(top_10_name, markers)
# volcano plot
options(ggrepel.max.overlaps = Inf)
olink_volcano_plot(p.val_tbl = ttest_results,
                   x_lab = 'Variant_vs_Healthy',
                   olinkid_list = labels)
cat("top 10 proteins and IL-6, IL-10 and PDCD1 are labeld")

```