---
title: "scRNA-seq PTCRA"
author: "Chao Di, dic@chop.edu"
output: 
  html_document:
    number_sections: true
    code_folding: hide
---

```{r, setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, 
                      warning = FALSE, 
                      tidy = TRUE,
                      fig.width = 10, 
                      tidy.opts = list(width.cutoff = 95))
 
```

# Abstract  
PTCRA patients in groups – PTCRA A, PTCRA B, PTCRA C and PTCRA D. The way of grouping patients may bring batch effect, which is difficult to tell. Need to find differentials within three patient/case - control groups.  

```{r, include=FALSE}
# library(scRNAseq)
library(scater)
library(SingleR)
library(celldex)
library(Seurat)
library(RColorBrewer)
library(sctransform)
library(glmGamPoi)
library(patchwork)
# library(EnhancedVolcano)
library(tidyverse)
library(reshape2)
```

# Prepare data
## Load Seurat object including all samples from Sam  
```{r, eval=FALSE}
# load Seurat obj
all.ptcra <- readRDS("ptcra_all_samples.rds") # original obj
Idents(all.ptcra) <- "sample"
```

## read meta data
```{r}
# read metadata
sample_info <- read.csv("sample_info.txt", sep="\t")
sample_info
```

## Add metadata info to obj 
last four columns: case_control, Condition, Age_group, Library, 'case_control' assignment will be replaced in the object  
```{r, eval = FALSE}
metadata_update <- left_join(all.ptcra@meta.data, sample_info %>% select(c("PID", "case_control", "Condition", "Age_group", "Library")), by = c("sample" = "PID"))
rownames(metadata_update) <- rownames(all.ptcra@meta.data)

all.ptcra <- AddMetaData(object = all.ptcra, metadata = metadata_update %>% select(c("case_control.y", "Condition", "Age_group", "Library")), col.name=c("case_control", "Condition", "Age_group", "Library"))
```

# Overiew of the data - Basic quality control and filtering  
Meta.data is the most important field for next steps. It can be acessed using both @ and [[]] operators. Right now it has 3 fields per celL: dataset ID, number of UMI reads detected per cell (nCount_RNA), and the number of expressed (detected) genes per same cell (nFeature_RNA).  

## Load "all.ptcra.rds"
Read filtered Seurat obj instead of re-run the following steps. 
```{r}
all.ptcra <- readRDS("all.ptcra.rds")
```

## Add percentage of counts for mitochondrial genes and add cell cycle scores   
```{r, eval = FALSE}
all.ptcra[["percent.mt"]] <- PercentageFeatureSet(all.ptcra, pattern = "^MT-")

s.genes <- cc.genes.updated.2019$s.genes
g2m.genes <- cc.genes.updated.2019$g2m.genes

all.ptcra <- CellCycleScoring(all.ptcra , s.features = s.genes, g2m.features = g2m.genes)
table(all.ptcra [[]]$Phase)
```

## Filter cells 
Filter cells that have unique feature counts over 2,500 (doublets or multiplets) or less than 200 and cells that have >5% mitochondrial counts  

```{r, eval=FALSE}
all.ptcra <- subset(all.ptcra, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
all.ptcra
```

## Remove TCR segment genes
```{r, eval=FALSE}
genes_to_remove <- read.table("PTCRA_genes_to_remove.txt", header=TRUE)
counts <- GetAssayData(all.ptcra, assay = "RNA")
counts <- counts[-(which(rownames(counts) %in% genes_to_remove$genes_to_remove)),]
all.ptcra <- subset(all.ptcra, features = rownames(counts))
```

## Save filtered obj
```{r, eval=FALSE}
saveRDS(all.ptcra, file="all.ptcra.rds")
```

## metadata information  
```{r}
all.ptcra
meta <- all.ptcra@meta.data
dim(meta)
summary(meta$nCount_RNA)
summary(meta$nFeature_RNA)
```

## Visualize QC metrics as a violin plot   
```{r, fig.width=10, fig.height=19}
VlnPlot(all.ptcra, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 1)
```

## Visualize feature-feature relationships 
```{r, fig.width=10, fig.height=8}
plot1 <- FeatureScatter(all.ptcra, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(all.ptcra, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

# Data Integration  

## Load object `all.ptcra.list`
Load `all.ptcra.list` instead of running the following chunks (set eval=FALSE) again when knitting.  
```{r, eval = FALSE}
all.ptcra.list <- readRDS("all.ptcra.list.rds")
```

## Split object and SCTransform normalization
Split object to gorups of patients and controls, normalize and identify variable features for each dataset independently. 

```{r, eval = FALSE}
all.ptcra.list <- SplitObject(all.ptcra, split.by = "case_control")

for (i in 1:length(all.ptcra.list)) {
    all.ptcra.list[[i]] <- SCTransform(all.ptcra.list[[i]], method = "glmGamPoi", vars.to.regress = c("percent.mt","S.Score","G2M.Score"), verbose = FALSE)
}
```

## Select features and prepare integration 
Select features that are repeatedly variable across datasets for integration, and run `PrepSCTIntegration`, which ensures that all necessary Pearson residuals have been calculated.  

```{r, eval = FALSE}
all.ptcra.features <- SelectIntegrationFeatures(object.list = all.ptcra.list, nfeatures = 3000)

all.ptcra.list <- PrepSCTIntegration(object.list = all.ptcra.list, anchor.features = all.ptcra.features)
```

## Run PCA on each object 
Run PCA on each object in the list, which is required for running the reciprocal PCA in `FindIntegrationAnchors`  

```{r, eval = FALSE}
for (i in 1:length(all.ptcra.list)) {
    all.ptcra.list[[i]] <- RunPCA(all.ptcra.list[[i]], features = all.ptcra.features)
}
```
## Save object `all.ptcra.list`
Save `all.ptcra.list` before extract specific datasets.  
```{r, eval = FALSE}
saveRDS(all.ptcra.list, file="all.ptcra.list.rds")
```

## Load objects `ptcra.integrated1`, `ptcra.integrated2`, and `ptcra.integrated3`
Load integrated objects instead of running the following chunks (set eval=FALSE) again when knitting.  
```{r}
ptcra.integrated1 <- readRDS("ptcra.integrated1.rds")
ptcra.integrated2 <- readRDS("ptcra.integrated2.rds")
ptcra.integrated3 <- readRDS("ptcra.integrated3.rds")
```

## Identify anchors and integrate the datasets
Make sure to set normalization.method = **"SCT"**, and reduction = **"rpca"** for reciprocal PCA 
```{r, eval = FALSE}
## Integrate all objects together, run all combinations of the objects. Time consuming.
# all.ptcra.anchors <- FindIntegrationAnchors(object.list = all.ptcra.list, normalization.method = "SCT", anchor.features = all.ptcra.features, reduction = "rpca")
# all.ptcra.integrated <- IntegrateData(anchorset = all.ptcra.anchors, normalization.method = "SCT")

ptcra.list1 <- c(all.ptcra.list$control1, all.ptcra.list$case1)
names(ptcra.list1) <- c("control1", "case1")
ptcra.anchors1 <- FindIntegrationAnchors(object.list = ptcra.list1, normalization.method = "SCT", anchor.features = all.ptcra.features, reduction = "rpca")
ptcra.integrated1 <- IntegrateData(anchorset = ptcra.anchors1, normalization.method = "SCT")

ptcra.list2 <- c(all.ptcra.list$control2, all.ptcra.list$case2)
names(ptcra.list2) <- c("control2", "case2")
ptcra.anchors2 <- FindIntegrationAnchors(object.list = ptcra.list2, normalization.method = "SCT", anchor.features = all.ptcra.features, reduction = "rpca")
ptcra.integrated2 <- IntegrateData(anchorset = ptcra.anchors2, normalization.method = "SCT")

ptcra.list3 <- c(all.ptcra.list$control3, all.ptcra.list$case3)
names(ptcra.list3) <- c("control3", "case3")
ptcra.anchors3 <- FindIntegrationAnchors(object.list = ptcra.list3, normalization.method = "SCT", anchor.features = all.ptcra.features, reduction = "rpca")
ptcra.integrated3 <- IntegrateData(anchorset = ptcra.anchors3, normalization.method = "SCT")

```

# Proceed with downstream analysis on the integrated dataset {.tabset}
## control1 and case1 integrated  
### PCA and Clustering  
```{r, eval = FALSE}
ptcra.integrated1 <- RunPCA(ptcra.integrated1)
ptcra.integrated1 <- RunUMAP(ptcra.integrated1, reduction = "pca", dims = 1:30)
ptcra.integrated1 <- FindNeighbors(ptcra.integrated1, reduction = "pca", dims = 1:30)
ptcra.integrated1 <- FindClusters(ptcra.integrated1, resolution = 0.5)
```

### Visualization  
```{r, fig.height=9, fig.width=16}
DefaultAssay(ptcra.integrated1) <- "integrated"

p1 <- DimPlot(ptcra.integrated1, reduction = "umap", group.by = "case_control") + theme(legend.position = "top")
p2 <- DimPlot(ptcra.integrated1, reduction = "umap", label = TRUE, repel = TRUE) + NoLegend()
p1 + p2
DimPlot(ptcra.integrated1, reduction = "umap", label = FALSE, split.by = "case_control") + theme(legend.position = "top")
```

### Cell type annotation using SingleR
Get reference datasets from `celldex` package. Note that there are two cell type assignments, `label.main` and `label.fine`. Convert our Seurat object to single cell experiment (SCE).   
```{r, eval = FALSE}
# use MonacoImmuneData as reference, can also use HumanPrimaryCellAtlasData or DatabaseImmuneCellExpressionData
monaco.ref <- celldex::MonacoImmuneData()

DefaultAssay(ptcra.integrated1) <- "RNA"
ptcra_sce1 <- as.SingleCellExperiment(DietSeurat(ptcra.integrated1))
ptcra_sce1
```
Run `SingleR`  

```{r, eval = FALSE}
ptcra_sce1.monaco.main <- SingleR(test = ptcra_sce1, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.main)
ptcra_sce1.monaco.fine <- SingleR(test = ptcra_sce1, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.fine)
```

Add the annotations to the Seurat object metadata (or replace previous labels).    
```{r, eval = FALSE}
ptcra.integrated1@meta.data$SingleR.labels_Monaco <- ptcra_sce1.monaco.fine$pruned.labels
```

### Compare the number of cells in each cell type
```{r}
ptcra.integrated1$celltype <- Idents(ptcra.integrated1)
table(ptcra.integrated1@meta.data$celltype, ptcra.integrated1@meta.data$Condition)

```

### plot comparisons
```{r, fig.height=8}
plot_integrated_celltypes = function (srat) { 
  count_table <- table(srat@meta.data$celltype, srat@meta.data$Condition)
  count_mtx   <- as.data.frame.matrix(count_table)
  count_mtx$celltype <- rownames(count_mtx)
  melt_mtx <- melt(count_mtx)
  

  celltype_size <- aggregate(value ~ celltype, data = melt_mtx, FUN = sum)
  celltype_size <- celltype_size %>% dplyr::arrange(value)
  celltype_size$celltype <- factor(celltype_size$celltype, levels=celltype_size$celltype)
  melt_mtx$celltype <- factor(melt_mtx$celltype, levels=celltype_size$celltype)
  colnames(melt_mtx)[2] <- "dataset"
  
  
  p1 <- ggplot(celltype_size, aes(y= celltype, x = value)) + geom_bar(position="dodge", stat="identity",fill = "grey60") + 
    theme_bw() + scale_x_log10() + xlab("Cells per celltype, log10 scale") + ylab("")
  p2 <- ggplot(melt_mtx,aes(x=celltype,y=value,fill=dataset)) + 
    geom_bar(position="fill", stat="identity") + theme_bw() + coord_flip() + 
    scale_fill_brewer(palette = "Set2") +
    ylab("Fraction of cells in each dataset") + xlab("celltype number") + theme(legend.position="top")
  
  p2 + p1 + plot_layout(widths = c(3,1))
}

plot_integrated_celltypes(ptcra.integrated1)
```


### Visualize the fine-grained annotations
```{r, fig.height=7, fig.width=14}
# Summary of general cell type annotations. 
# table(ptcra_sce1.monaco.main$pruned.labels)
# table(ptcra_sce1.monaco.fine$pruned.labels)
# remove NA labeled samples
ptcra.integrated1 <- ptcra.integrated1[,!is.na(ptcra.integrated1$SingleR.labels_Monaco)]

ptcra.integrated1 <- SetIdent(ptcra.integrated1, value = "SingleR.labels_Monaco")
DimPlot(ptcra.integrated1, reduction = "umap", label = TRUE, label.size=3, repel = TRUE, split.by = "case_control") + theme(legend.position = "top")
```

Plot some markers.  
```{r}
FeaturePlot(ptcra.integrated1,c("CCR7", "SELL","PTPRC", "IL7R","IL2RG","CD27", "CD79A", "FCGR3A")) 
```

Save object `ptcra.integrated1`.    
```{r, eval= FALSE}
saveRDS(ptcra.integrated1, file="ptcra.integrated1.rds")
```

### Identify differential expressed genes between patients vs controls for all cells
We’ve aligned the patient and control cells. We choose **all cells** for differential expression analysis between patients and controls.
We create a column in the meta.data slot to hold both the cell type and case_control information and switch the current ident to that column. Then we use `FindMarkers()` to find the genes that are different between patient and control. 

#### Load previous results
```{r}
set1.response <- readRDS("PTCRA_vs_HC_DE_case1.rds")

head(set1.response %>% arrange(desc(avg_log2FC)), 5)
head(set1.response %>% arrange(avg_log2FC), 5)
```

#### Finding differential expression markers 
```{r, eval=FALSE}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

ptcra.integrated1$celltype.case_control <- paste(ptcra.integrated1$celltype, ptcra.integrated1$case_control, sep = "_")

Idents(ptcra.integrated1) <- "case_control"
set1.response <- FindMarkers(ptcra.integrated1, ident.1 = "case1", ident.2 = "control1", min.pct = 0.25)
set1.response <- set1.response %>% filter(p_val_adj<0.05)

```
#### Save DE results to table
```{r, eval=FALSE}
write.table(data.frame("GeneSymbol"=rownames(set1.response), set1.response), file="PTCRA_vs_HC_DE_case1.tsv", sep="\t", quote=FALSE, col.names = TRUE, row.names=FALSE )
saveRDS(set1.response, file="PTCRA_vs_HC_DE_case1.rds")
```


#### Visualization
```{r, fig.height=12}
FeaturePlot(ptcra.integrated1, features = c("RPS26", "CCL5", "CD8A"), split.by = "case_control", max.cutoff = 3,
    cols = c("grey", "red"))

plots <- VlnPlot(ptcra.integrated1, features = c("RPS26", "CCL5", "CD8A"), split.by = "case_control", group.by = "celltype",
    pt.size = 0, combine = FALSE)
wrap_plots(plots = plots, ncol = 1)
```
##### Heatmap 
```{r}

top20markers <- rbind(head(set1.response %>% arrange(desc(avg_log2FC)), 25), head(set1.response %>% arrange(avg_log2FC), 25))

count_table <- table(ptcra.integrated1@meta.data$celltype, ptcra.integrated1@meta.data$Condition)
count_mtx   <- as.data.frame.matrix(count_table)
count_mtx$celltype <- rownames(count_mtx)
melt_mtx <- melt(count_mtx)
celltype_size <- aggregate(value ~ celltype, data = melt_mtx, FUN = sum)
top_cells <- celltype_size %>% arrange(desc(value)) %>% filter(value>=200)
  
Idents(ptcra.integrated1) <- "celltype"

pdf("case1_vs_control1_heatmap.pdf", 8,8)
# all cells together
p1 <- DoHeatmap(ptcra.integrated1, features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = "all cells", theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p1)
for (cell in top_cells$celltype) {
p <- DoHeatmap(subset(ptcra.integrated1, idents=cell), features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = cell, theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p)
}
dev.off()
```

## control2 and case2 integrated  
### PCA and Clustering 
```{r, eval= FALSE}
ptcra.integrated2 <- RunPCA(ptcra.integrated2)
ptcra.integrated2 <- RunUMAP(ptcra.integrated2, reduction = "pca", dims = 1:30)
ptcra.integrated2 <- FindNeighbors(ptcra.integrated2, reduction = "pca", dims = 1:30)
ptcra.integrated2 <- FindClusters(ptcra.integrated2, resolution = 0.5)

```

### Visualization  
```{r, fig.height=9, fig.width=16}
DefaultAssay(ptcra.integrated2) <- "integrated"
p1 <- DimPlot(ptcra.integrated2, reduction = "umap", group.by = "case_control") + theme(legend.position = "top")
p2 <- DimPlot(ptcra.integrated2, reduction = "umap", label = TRUE, repel = TRUE) + theme(legend.position = "top")
p1 + p2
DimPlot(ptcra.integrated2, reduction = "umap", split.by = "case_control") + theme(legend.position = "top")
```

### Cell type annotation using SingleR

Get reference datasets from `celldex` package. Note that there are two cell type assignments, `label.main` and `label.fine`. Convert our Seurat object to single cell experiment (SCE).   
```{r, eval = FALSE}
# use MonacoImmuneData as reference, can also use HumanPrimaryCellAtlasData or DatabaseImmuneCellExpressionData
monaco.ref <- celldex::MonacoImmuneData()

DefaultAssay(ptcra.integrated2) <- "RNA"
ptcra_sce2 <- as.SingleCellExperiment(DietSeurat(ptcra.integrated2))
ptcra_sce2
```

Run `SingleR`  
```{r, eval = FALSE}
ptcra_sce2.monaco.main <- SingleR(test = ptcra_sce2, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.main)
ptcra_sce2.monaco.fine <- SingleR(test = ptcra_sce2, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.fine)
```

Add the annotations to the Seurat object metadata (or replace previous labels).    
```{r, eval = FALSE}
ptcra.integrated2@meta.data$SingleR.labels_Monaco <- ptcra_sce2.monaco.fine$pruned.labels
```

### Compare the number of cells in each cell type
```{r}
ptcra.integrated2$celltype <- Idents(ptcra.integrated2)
table(ptcra.integrated2@meta.data$celltype, ptcra.integrated2@meta.data$Condition)

```

### plot comparisons
```{r, fig.height=8}
plot_integrated_celltypes(ptcra.integrated2)
```

### Visualize the fine-grained annotations
```{r, fig.height=8, fig.width=14}
# Summary of general cell type annotations
# table(ptcra_sce2.monaco.main$pruned.labels)
# table(ptcra_sce2.monaco.fine$pruned.labels)

ptcra.integrated2 <- ptcra.integrated2[,!is.na(ptcra.integrated2$SingleR.labels_Monaco)]
ptcra.integrated2 <- SetIdent(ptcra.integrated2, value = "SingleR.labels_Monaco")
table(ptcra.integrated2$SingleR.labels_Monaco)

DimPlot(ptcra.integrated2, label = TRUE , repel = TRUE, label.size = 3, split.by="case_control") + theme(legend.position = "top")
```

Plot some markers.  
```{r}
FeaturePlot(ptcra.integrated2,c("CCR7", "SELL","PTPRC", "IL7R","IL2RG","CD27", "CD79A", "FCGR3A")) 
```
Save object.  
```{r, eval = FALSE}
saveRDS(ptcra.integrated2, file="ptcra.integrated2.rds")
```

### Identify differential expressed genes between patients vs controls for specific cell types 
We’ve aligned the patient and control cells. We choose **Naive CD8 T cells** populations for differential expression analysis between patients and controls.
We create a column in the meta.data slot to hold both the cell type and case_control information and switch the current ident to that column. Then we use `FindMarkers()` to find the genes that are different between patient and control "Naive CD8 T cells". 

#### Load previous results
```{r}
set2.response <- readRDS("PTCRA_vs_HC_DE_case2.rds")

head(set2.response %>% arrange(desc(avg_log2FC)), 5)
head(set2.response %>% arrange(avg_log2FC), 5)
```

#### Finding differential expression markers 
```{r, eval=FALSE}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

ptcra.integrated2$celltype.case_control <- paste(ptcra.integrated2$celltype, ptcra.integrated2$case_control, sep = "_")

Idents(ptcra.integrated2) <- "case_control"
set2.response <- FindMarkers(ptcra.integrated2, ident.1 = "case2", ident.2 = "control2", min.pct = 0.25)
set2.response <- set2.response %>% filter(p_val_adj<0.05)
```
#### Save DE results to table
```{r, eval=FALSE}
write.table(data.frame("GeneSymbol"=rownames(set2.response), set2.response), file="PTCRA_vs_HC_DE_case2.tsv", sep="\t", quote=FALSE, col.names = TRUE, row.names=FALSE )
saveRDS(set2.response, file="PTCRA_vs_HC_DE_case2.rds")
```


#### Visualization  
```{r, fig.height=12}
FeaturePlot(ptcra.integrated2, features = c("CRIP1", "ANXA1", "RPS26"), split.by = "case_control", max.cutoff = 3,
    cols = c("grey", "red"))

plots <- VlnPlot(ptcra.integrated2, features = c("CRIP1", "ANXA1", "RPS26"), split.by = "case_control", group.by = "celltype",
    pt.size = 0, combine = FALSE)
wrap_plots(plots = plots, ncol = 1)
```

##### Heatmap 
```{r}

top20markers <- rbind(head(set2.response %>% arrange(desc(avg_log2FC)), 25), head(set2.response %>% arrange(avg_log2FC), 25))

count_table <- table(ptcra.integrated2@meta.data$celltype, ptcra.integrated2@meta.data$Condition)
count_mtx   <- as.data.frame.matrix(count_table)
count_mtx$celltype <- rownames(count_mtx)
melt_mtx <- melt(count_mtx)
celltype_size <- aggregate(value ~ celltype, data = melt_mtx, FUN = sum)
top_cells <- celltype_size %>% arrange(desc(value)) %>% filter(value>=200)
  
Idents(ptcra.integrated2) <- "celltype"

pdf("case2_vs_control2_heatmap.pdf", 8,8)
# all cells together
p1 <- DoHeatmap(ptcra.integrated2, features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = "all cells", theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p1)
for (cell in top_cells$celltype) {
p <- DoHeatmap(subset(ptcra.integrated2, idents=cell), features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = cell, theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p)
}
dev.off()
```

## control3 and case3 integrated  
### PCA and Clustering 
```{r, eval = FALSE}
ptcra.integrated3 <- RunPCA(ptcra.integrated3)
ptcra.integrated3 <- RunUMAP(ptcra.integrated3, reduction = "pca", dims = 1:30)
ptcra.integrated3 <- FindNeighbors(ptcra.integrated3, reduction = "pca", dims = 1:30)
ptcra.integrated3 <- FindClusters(ptcra.integrated3, resolution = 0.5)
```

### Visualization  
```{r, fig.height=9, fig.width=16}
DefaultAssay(ptcra.integrated3) <- "integrated"
options(ggrepel.max.overlaps = Inf)
p1 <- DimPlot(ptcra.integrated3, reduction = "umap", group.by = "case_control") + theme(legend.position = "top")
p2 <- DimPlot(ptcra.integrated3, reduction = "umap", label = TRUE, repel = TRUE) + theme(legend.position = "top")
p1 + p2
DimPlot(ptcra.integrated3, reduction = "umap", split.by = "case_control") + theme(legend.position = "top")
```


### Cell type annotation using SingleR
Get reference datasets from `celldex` package. Note that there are two cell type assignments, `label.main` and `label.fine`. Convert our Seurat object to single cell experiment (SCE).   
```{r, eval = FALSE}
# use MonacoImmuneData as reference, can also use HumanPrimaryCellAtlasData or DatabaseImmuneCellExpressionData
monaco.ref <- celldex::MonacoImmuneData()

DefaultAssay(ptcra.integrated3) <- "RNA"
ptcra_sce3 <- as.SingleCellExperiment(DietSeurat(ptcra.integrated3))
ptcra_sce3
```

Run `SingleR`  
```{r, eval = FALSE}
ptcra_sce3.monaco.main <- SingleR(test = ptcra_sce3, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.main)
ptcra_sce3.monaco.fine <- SingleR(test = ptcra_sce3, assay.type.test = 1, ref = monaco.ref, labels = monaco.ref$label.fine)
```

Add the annotations to the Seurat object metadata (or replace previous labels).    
```{r, eval = FALSE}
ptcra.integrated3@meta.data$SingleR.labels_Monaco <- ptcra_sce3.monaco.fine$pruned.labels
```

### Compare the number of cells in each cell type
```{r}
ptcra.integrated3$celltype <- Idents(ptcra.integrated3)
table(ptcra.integrated3@meta.data$celltype, ptcra.integrated3@meta.data$Condition)

```

### plot comparisons
```{r, fig.height=8}
plot_integrated_celltypes(ptcra.integrated3)
```

### Visualize the fine-grained annotations
```{r, fig.height=8, fig.width=15}
# Summary of general cell type annotations.  
# table(ptcra_sce3.monaco.main$pruned.labels)
# table(ptcra_sce3.monaco.fine$pruned.labels)
# remove NA samples 
ptcra.integrated3 <- ptcra.integrated3[,!is.na(ptcra.integrated3$SingleR.labels_Monaco)]
table(ptcra.integrated3$SingleR.labels_Monaco)

ptcra.integrated3 <- SetIdent(ptcra.integrated3, value = "SingleR.labels_Monaco")
DimPlot(ptcra.integrated3, label = TRUE , repel = TRUE, label.size = 3, split.by="case_control") + theme(legend.position = "top")
```

Plot some markers.  
```{r}
FeaturePlot(ptcra.integrated3,c("CCR7", "SELL","PTPRC", "IL7R","IL2RG","CD27", "CD79A", "FCGR3A", "PPBP")) 
```
Save object.  
```{r, eval = FALSE}
saveRDS(ptcra.integrated3, file="ptcra.integrated3.rds")
```

### Identify differential expressed genes between patients vs controls for specific cell types 
We’ve aligned the patient and control cells. We choose **Naive CD8 T cells** populations for differential expression analysis between patients and controls.
We create a column in the meta.data slot to hold both the cell type and case_control information and switch the current ident to that column. Then we use `FindMarkers()` to find the genes that are different between patient and control "Naive CD8 T cells". 

#### Load previous results
```{r}
set3.response <- readRDS("PTCRA_vs_HC_DE_case3.rds")

head(set3.response %>% arrange(desc(avg_log2FC)), 5)
head(set3.response %>% arrange(avg_log2FC), 5)
```

#### Finding differential expression markers 

```{r, eval=FALSE}
library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())

ptcra.integrated3$celltype <- Idents(ptcra.integrated3)
ptcra.integrated3$celltype.case_control <- paste(ptcra.integrated3$celltype, ptcra.integrated3$case_control, sep = "_")

Idents(ptcra.integrated3) <- "case_control"
set3.response <- FindMarkers(ptcra.integrated3, ident.1 = "case3", ident.2 = "control3", min.pct = 0.25)
set3.response <- set3.response %>% filter(p_val_adj<0.05)

```

#### Save DE results to table
```{r, eval=FALSE}
write.table(data.frame("GeneSymbol"=rownames(set3.response), set3.response), file="PTCRA_vs_HC_DE_case3.tsv", sep="\t", quote=FALSE, col.names = TRUE, row.names=FALSE )
saveRDS(set3.response, file="PTCRA_vs_HC_DE_case3.rds")
```

#### Visualization   
```{r, fig.height=12}
FeaturePlot(ptcra.integrated3, features = c("OASL", "GZMK", "GAS5"), split.by = "case_control", max.cutoff = 3,
    cols = c("grey", "red"))

plots <- VlnPlot(ptcra.integrated3, features = c("OASL", "GZMK", "GAS5"), split.by = "case_control", group.by = "celltype",
    pt.size = 0, combine = FALSE)
wrap_plots(plots = plots, ncol = 1)
```

##### Heatmap 
```{r}

top20markers <- rbind(head(set3.response %>% arrange(desc(avg_log2FC)), 25), head(set3.response %>% arrange(avg_log2FC), 25))

count_table <- table(ptcra.integrated3@meta.data$celltype, ptcra.integrated3@meta.data$Condition)
count_mtx   <- as.data.frame.matrix(count_table)
count_mtx$celltype <- rownames(count_mtx)
melt_mtx <- melt(count_mtx)
celltype_size <- aggregate(value ~ celltype, data = melt_mtx, FUN = sum)
top_cells <- celltype_size %>% arrange(desc(value)) %>% filter(value>=200)
  
Idents(ptcra.integrated3) <- "celltype"

pdf("case3_vs_control3_heatmap.pdf", 8,8)
# all cells together
p1 <- DoHeatmap(ptcra.integrated3, features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = "all cells", theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p1)
for (cell in top_cells$celltype) {
p <- DoHeatmap(subset(ptcra.integrated3, idents=cell), features = rownames(top20markers),
          group.by = c("case_control"),
          assay = "SCT",
          label = FALSE) + plot_annotation(title = cell, theme = theme(plot.title = element_text(size = 18, hjust = 0.5)))
print(p)
}
dev.off()
```

